---
title: "DSA2101 Group Project"
output: html_document
date: "2025-04-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Main question: How do returning participants influence a country's performance in IMO, and has this relationship changed over time?

## Library

```{r}
library(tidytuesdayR)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Importing Data

```{r tuesdata}
tuesdata <- tidytuesdayR::tt_load('2024-09-24')
```

### Country Results

cleaning here first for total

```{r country_results_df}
country_results_df <- tuesdata$country_results_df

country_results_df %>%
  summarise(across(everything(), ~ sum(is.na(.))))
```

### Individual Results

cleaning here first for total

```{r individual_results_df}
individual_results_df <- tuesdata$individual_results_df

individual_results_df %>% 
  summarise(across(everything(), ~ sum(is.na(.))))
```

### Timeline

```{r timeline_df}
timeline_df <- tuesdata$timeline_df

na_timeline <- timeline_df %>% 
  summarise(across(everything(), ~ sum(is.na(.))))
```

## Data cleaning

```{r removing na from country_results_df}
country_results_df <- country_results_df %>%
  mutate(team_size_male = ifelse(is.na(team_size_male), 0, team_size_male),
         team_size_female = ifelse(is.na(team_size_female), 0, team_size_female),
         team_size_all = ifelse(is.na(team_size_all), 0, team_size_all),
         p1 = ifelse(is.na(p1), 0, p1),
         p2 = ifelse(is.na(p2), 0, p2),
         p3 = ifelse(is.na(p3), 0, p3),
         p4 = ifelse(is.na(p4), 0, p4),
         p5 = ifelse(is.na(p5), 0, p5),
         p6 = ifelse(is.na(p6), 0, p6),
         p7 = ifelse(is.na(p7), 0, p7),
         awards_gold = ifelse(is.na(awards_gold), 0, awards_gold),
         awards_silver = ifelse(is.na(awards_silver), 0, awards_silver),
         awards_bronze = ifelse(is.na(awards_bronze), 0, awards_bronze),
         awards_honorable_mentions = ifelse(is.na(awards_honorable_mentions), 0, awards_honorable_mentions),
         leader = ifelse(is.na(leader), "None", leader),
         deputy_leader = ifelse(is.na(deputy_leader), "None", deputy_leader))
```

```{r removing na from individual_results_df}
individual_results_df <- individual_results_df %>%
  mutate(p1 = ifelse(is.na(p1), 0, p1),
         p2 = ifelse(is.na(p2), 0, p2),
         p3 = ifelse(is.na(p3), 0, p3),
         p4 = ifelse(is.na(p4), 0, p4),
         p5 = ifelse(is.na(p5), 0, p5),
         p6 = ifelse(is.na(p6), 0, p6),
         p7 = ifelse(is.na(p7), 0, p7),
         individual_rank = ifelse(is.na(individual_rank), 0, individual_rank),
         award = ifelse(is.na(award), "None", award))

```

```{r removing na from timeline_df}
timeline_df <- timeline_df %>%
  mutate(female_contestant = ifelse(is.na(female_contestant), 0, female_contestant))
```

## Question 1

Qn 1: How has the proportion of returning participants evolved over the years for each countries?

Since there are a lot of countries, we scope into the top 5 countries in the latest year, and look into their past performances in the IMO.

### Top 5 Coutries

```{r top_5_countries_in_2024}
top_5_countries_in_2024 <- country_results_df %>%
  filter(year == 2024) %>%
  mutate(total = rowSums(across(p1:p7))) %>%
  arrange(desc(total), 
          desc(awards_gold), 
          desc(awards_silver), 
          desc(awards_bronze), 
          desc(awards_honorable_mentions)) %>%
  head(5) %>%
  pull(country)
```

### Check any missing record after 2000 year

For simplicity, we only look at the data that is after 2000 year for those 5 countries

```{r Check missing record}
top_5_countries_results_df <- country_results_df %>%
  filter(country %in% top_5_countries_in_2024, year >= 2000) %>%
  group_by(country) %>%
  arrange(country) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  glimpse()
```

Turns out that there is one missing record for India, but it should not be problem

### Get all the participant and results of those countries

Get to know which participant are returning participant

```{r}
# I marked the participant with 1 if he/she got more than 1 appearance in the df
individual_return <- individual_results_df %>%
  filter(country %in% top_5_countries_in_2024) %>%
  arrange(contestant, year) %>%
  group_by(contestant) %>%
  mutate(appearance = row_number()) %>%
  ungroup() %>%
  mutate(returning = if_else(appearance > 1, 1, 0)) %>%
  filter(year >= 2000)
```

```{r}
top5_country_results <- country_results_df %>%
  filter(country %in% top_5_countries_in_2024, year >= 2000)
```

### Get the proportion

```{r Calculating the proportion}
# since returning participant have 1 value of returning, so summing up them should give the total of returning participant, maybe
returning_summary <- individual_return %>%
  group_by(country, year) %>%
  summarize(returning_count = sum(returning), .groups = "drop")

returning_summary <- returning_summary %>%
  left_join(
    top5_country_results %>% select(country, year, team_size_all), 
    by = c("country", "year")
  ) %>%
  mutate(proportion_returning = returning_count / team_size_all)
```

### Plotting

```{r}
ggplot(returning_summary, aes(x = year, y = proportion_returning, color = country)) +
  geom_line() +
  labs(
    title = "Proportion of Returning Participants Over Time",
    x = "Year",
    y = "Proportion of Returning Participants"
  )
```

## Question 2

Qn 2: Do returning participants generally achieve higher score compared to first-time participants

### Aggregating the data

```{r}
# Divide into two, one is the participant who only join one time, other is returning participant
paired_individual <- top5_individual %>% 
  group_by(contestant) %>%
  filter(n() >= 2) %>%  # keep only those with 2+ appearances
  ungroup() %>%
  mutate(participant_type = if_else(appearance == 1, "Join only one time", "Returning participant"))
```

### Plotting

```{r}
ggplot(paired_individual, aes(x = participant_type, y = total, fill = participant_type)) +
  geom_boxplot() +
  labs(
    title = "Score Comparison: First Attempt vs. Returning Attempt",
    x = "Participant Type",
    y = "Total Score"
  ) +
  theme_minimal()
```

## Question 3

Qn 3: Does a higher proportion of returning participants on a team correlate with improved overall team performances?

### Adding the column (total) for all country result

```{r}
team_performance <- top5_country_results %>%
  mutate(total = p1 + p2 + p3 + p4 + p5 + p6 + p7) %>%
  select(country, year, total)
```

### Join two tables to get proportion and total points together

```{r}
final_team_data <- returning_summary %>%
  left_join(team_performance, by = c("country", "year"))
```

### Plotting

```{r}
# I used additional line to see the trend, looks like good bc we see improvement when the propotion goes up
ggplot(final_team_data, aes(x = proportion_returning, y = total, color = country)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Team Performance vs. Proportion of Returning Participants",
    x = "Proportion of Returning Participants",
    y = "Total Team Score"
  ) +
  theme_minimal()
```



